True versus apparent shapes of bow shocks

### 2018-03-02 papers/Papers1234/app-rcurv-empirical.tex
Make table of the M42 bow fits

Consider a set of \(N\) points on the plane of the sky, with Cartesian
coordinates \(\bm{r}_k = (x_k, y_k)\) for \(k = 1 \dots N\).  We wish
to estimate the radius of curvature of the smooth curve that the set
of points is presumed to be sampled from.  To do this, we fit a circle
to the points as follows.  The circle is defined by its center,
\(\bm{r}_{\C} = (x_{\C}, y_{\C})\), and radius, \(R_{\C}\).  For a
given circle, we define a mean radius of the set of points from the
circle center:
\begin{equation}
  \label{eq:rcurv-mean-radius}
  \bar{R}_\C(x_{\C}, y_{\C}) = \frac{1}{N}  \sum_{k = 1}^{N} \Abs{\bm{r}_k - \bm{r}_{\C}}  \ .
\end{equation}
We then optimize
% \footnote{%
%   For example, using a Levenberg-Marquardt algorithm, such as
%   \texttt{astropy.modeling.fitting.LevMarLSQFitter} from the AstroPy
%   library \citep{Astropy-Collaboration:2013a}.} %
to find best-fit values \((x_{\C}^*, y_{\C}^*)\), which minimize the
objective function
\begin{equation}
  \label{eq:rcurv-objective-function}
  f(x_{\C}, y_{\C}) = \sum_{k = 1}^{N} \left(
    \abs{\bm{r}_k - \bm{r}_{\C}}  - \bar{R}_\C (x_{\C}, y_{\C}) \right)^2 \ .
\end{equation}
The best-fit radius of curvature is then given by
\(R_{\C}^* = \bar{R}_\C(x_{\C}^*, y_{\C}^*)\).

@@

Since there is no guarantee of symmetry about the axis \(\uvec{\xi}\),
the perpendicular radius will in general be different in the two wings
of the bow, with values \(R_{90}^+\) and \(R_{90}^-\).  These can be
estimated by defining
\begin{equation}
  \label{eq:rcurv-R-k}
  R_k = \Abs{\bm{r}_k - \bm{r}_0}
\end{equation}
and linearly interpolating between the points \((\theta_k, R_k)\) at
\(\theta = \pm \ang{90}\).

>>>>

Consider a set of \(N\) points on the plane of the sky,\footnote{%
  In the main body of the paper, the prime symbol (\('\)) is used to
  distinguish projected from ``true'' quantities.  In this Appendix,
  for simplicity, we omit the primes since \emph{all} quantities are
  projected.} %
with Cartesian coordinates \(\bm{r}_k = (x_k, y_k)\) for
\(k = 1 \dots N\).  We wish to estimate the radius of curvature of the
smooth curve that the set of points is presumed to be sampled from.
To do this, we fit a circle to the points as follows.  The circle is
defined by its center, \(\bm{r}_{\C} = (x_{\C}, y_{\C})\), and radius,
\(R_{\C}\).  For a given circle, we define a mean radius of the set of
points from the circle center:
\begin{equation}
  \label{eq:rcurv-mean-radius}
  \bar{R}_\C(x_{\C}, y_{\C}) = \frac{1}{N}  \sum_{k = 1}^{N} \Abs{\bm{r}_k - \bm{r}_{\C}}  \ .
\end{equation}
We then optimize
% \footnote{%
%   For example, using a Levenberg-Marquardt algorithm, such as
%   \texttt{astropy.modeling.fitting.LevMarLSQFitter} from the AstroPy
%   library \citep{Astropy-Collaboration:2013a}.} %
to find best-fit values \((x_{\C}^*, y_{\C}^*)\), which minimize the
objective function
\begin{equation}
  \label{eq:rcurv-objective-function}
  f(x_{\C}, y_{\C}) = \sum_{k = 1}^{N} \left(
    \abs{\bm{r}_k - \bm{r}_{\C}}  - \bar{R}_\C (x_{\C}, y_{\C}) \right)^2 \ .
\end{equation}
The best-fit radius of curvature is then given by
\(R_{\C}^* = \bar{R}_\C(x_{\C}^*, y_{\C}^*)\).

@@

Since there is no guarantee of symmetry about the axis \(\uvec{\xi}\),
the perpendicular radius will in general be different in the two wings
of the bow, with values \(R_{90+}\) and \(R_{90-}\).  These can be
estimated by defining
\begin{equation}
  \label{eq:rcurv-R-k}
  R_k = \Abs{\bm{r}_k - \bm{r}_0}
\end{equation}
and linearly interpolating between the points \((\theta_k, R_k)\) at
\(\theta = \pm \ang{90}\).

### 2018-03-02 papers/Papers1234/sec-conclusions.tex
Make table of the M42 bow fits

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figs/new-two-arc-multi-fig}
  \caption[]{Example empirical determination of planitude and alatude
    for observed bow shocks associated with proplyds in the Orion
    Nebula.  Color scale and contours show an \textit{HST} H\(\alpha\)
    image (ACS F658N filter, \citealp{Bally:2006a}) of M42~000-400
    (panels \textit{a--c}) and M42~069-601 (panels \textit{d--f}). Three
    different bows have been traced by hand on the image (red and
    white filled symbols): (\textit{a, d})~inner edge, (\textit{b,
      e})~ridge of maximum emission, and (\textit{c, f})~outer edge.
    For each panel, the dark-colored circle shows the initial fit to
    the full set of points (white and red), using the algorithm
    described in Appendix~\ref{app:rcurv-empirical}.  The center of
    curvature and derived axis are shown by a small filled circle and
    dashed line in the same color. Lighter colored circles show three
    subsequent iterations where the fit is restricted to points within
    \(\pm \Delta\theta = \ang{75}\) of the axis.  The subset of points used in
    the final iteration is marked in red.  The perpendicular radii for
    the final iteration are shown by dotted lines. In panels
    \textit{a, b, d--f} the iterations converge immediately, but in
    panel \textit{c} the iterations stably oscillate between two
    slightly different solutions. }
  \label{fig:000-400-fit}
\end{figure*}

\begin{figure*}
  \begin{tabular}{p{0.45\linewidth} p{0.45\linewidth}}
    (a) & (b) \\
    \includegraphics[width=\linewidth]{figs/000-400-planitude-alatude}
    & \includegraphics[width=\linewidth]{figs/069-601-planitude-alatude}
  \end{tabular}
  \caption[]{Location in the projected planitude--alatude plane of the
    converged circle fits to the M42~000-400 bow.  For each solution,
    the two values of the alatude \(\Lambda'\), corresponding to
    \(R_{90}^+\) and \(R_{90}^-\), are joined by a vertical
    line. Large symbols show the results from the fits shown in
    Figure~\ref{fig:000-400-fit}, while small symbols show results for
    fits using \(\Delta\theta = \ang{60}\) instead of \ang{75}.}
  \label{fig:000-400-planitude-alatude}
\end{figure*}

\begin{figure*}
  \begin{tabular}{p{0.45\linewidth} p{0.45\linewidth}}
    (a) & (b) \\
    \includegraphics[width=\linewidth]{figs/000-400-Pi-vs-Dtheta}
    & \includegraphics[width=\linewidth]{figs/069-601-Pi-vs-Dtheta}
  \end{tabular}
  \caption[]{Sensitivity of fitted planitude to the parameter
    \(\Delta\theta\), which controls how close a point must be to the axis in
    order to be included in the circle fit. }
  \label{fig:000-400-Delta-theta}
\end{figure*}

@@

As an example of measuring the projected planitude and alatude of a
real bow shock, we present an analysis of the H\(\alpha\) emission arc
associated with the proplyd M42~000-400 in the west of the Orion
Nebula \citep{Bally:2000a},\footnote{%
  The same object is listed as 4596-400 in some catalogs, such as
  \citet{Ricci:2008a}.  } %
which was used as an archetype of the bow shape in
Figure~\ref{fig:bow-terminology}.  To give an idea of the systematic
uncertainties and sensitivity to subjective choices, three different
tracings of the bow shape are considered (see
Fig.~\ref{fig:000-400-fit}): the peak of the emission arc (``ridge''),
and its inner and outer edges.  In all three
cases, we placed the points that define the bow by eye, using
SAOImage~DS9 in a similar fashion to in Figure~\ref{fig:meyer-trace},
and guided by the image contours.\footnote{%
  In the case of the ``ridge'' method at least, it is possible to
  automate this step, which we will discuss in detail in a following
  paper. } %
The principal hindrance to an accurate tracing of the bow shape in
this object is the strong and variable nebular background emission,
which particularly affects the far wing on the right hand side of the
image, where a bright background filament is superimposed on the bow.

We determine the planitude and alatude by fitting a circle to the
traced points within \(\pm \Delta\theta = \ang{75}\) bow axis, using the
iterative algorithm described in Appendix~\ref{app:rcurv-empirical}.
The fitted circle, when combined with the position of the central
source, yields the orientation of the bow axis, together with the apex
distance, \(R_0\), radius of curvature, \(R_\C\), and two
perpendicular radii (one for each wing), \(R_{90}^+\) and
\(R_{90}^-\).  These are all indicated on the panels of
Figure~\ref{fig:000-400-fit} by light-colored lines.\footnote{%
  For conciseness, we drop the prime symbol from the radii in this
  section and Appendix~\ref{app:rcurv-empirical}, even though all the
  quantities are projected ones.} %
The projected planitude and alatude then follow as \(\Pi = R_\C/R_0\),
\(\Lambda^+ = R_{90}^+/R_0\), \(\Lambda^- = R_{90}^-/R_0\), which are shown in Figure~\ref{fig:000-400-planitude-alatude}.

The planitude is found to have a moderate dependence on the choice of
\(\Delta\theta\), as shown in Figure~\ref{fig:000-400-Delta-theta}. The fact
that the radius of curvature is defined at a point (the projected
apex) might seem to argue for making \(\Delta\theta\) as small as possible, but
that would lead to circle fits that were extremely sensitive to the
exact positions of the few points included in the fit.  A reliable fit
requires 4 or more points, ideally spanning a total separation that is
a substantial fraction of \(R_\C\), which would argue for
\(\Delta\theta\) larger than about \(\Pi/2\) radians, or \ang{60}--\ang{90}.  On
the other hand, if \(\Delta\theta \ge \ang{90}\) then the planitude and alatude
would no longer be independent since the bow would be forced to lie on
the ``sphere'' line, \(\Lambda = (2 \Pi - 1)^{1/2}\) (see \S~\ref{sec:conic}).
Balancing these two concerns suggests an optimal
\(\Delta\theta = \ang{75}\), which is what is used in
Figure~\ref{fig:000-400-fit}, but we also show results for both
\ang{75} (thick lines) and \ang{60} (thin lines) in
Figure~\ref{fig:000-400-planitude-alatude}.

\(\Pi = 2.9 \pm 0.2 \pm 0.3\)

>>>>

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figs/new-two-arc-multi-fig}
  \caption[]{Example empirical determination of planitude and alatude
    for observed bow shocks associated with proplyds in the Orion
    Nebula (M42).  Color scale and contours show an \textit{HST}
    H\(\alpha\) image (ACS F658N filter, \citealp{Bally:2006a}) of
    M42~000-400 (panels \textit{a--c}) and M42~069-601 (panels
    \textit{d--f}).  The image scale and orientation are indicated on
    panel~\textit{a} and are the same for all panels.  Three different
    bows have been traced by eye on each object (red and white filled
    symbols): (\textit{a, d})~inner edge, (\textit{b, e})~ridge of
    maximum emission, and (\textit{c, f})~outer edge.  For each panel,
    the dark-colored circle shows the initial fit to the full set of
    points (white and red), using the algorithm described in
    Appendix~\ref{app:rcurv-empirical}.  The center of curvature and
    derived axis are shown by a small filled circle and dashed line in
    the same color. Lighter colored circles show three subsequent
    iterations where the fit is restricted to points within
    \(\pm \Delta\theta = \ang{75}\) of the axis.  The subset of points used in
    the final iteration is marked in red.  The perpendicular radii for
    the final iteration are shown by dotted lines. In panels
    \textit{a, b, d--f} the iterations converge immediately, but in
    panel \textit{c} the iterations stably oscillate between two
    slightly different solutions. }
  \label{fig:000-400-fit}
\end{figure*}

\begin{figure*}
  \begin{tabular}{p{0.47\linewidth} p{0.47\linewidth}}
    (a) & (b) \\
    \includegraphics[width=\linewidth]{figs/000-400-planitude-alatude}
    & \includegraphics[width=\linewidth]{figs/069-601-planitude-alatude}
  \end{tabular}
  \caption[]{Location in the projected planitude--alatude plane of the
    converged circle fits to the M42 bows: (\textit{a})~000-400,
    (\textit{b})~069-601. For each solution, the two values of the
    projected alatude, \(\Lambda_+'\) and \(\Lambda_-'\), corresponding to
    \(R_{90+}\) and \(R_{90-}\), are joined by a vertical line. Large
    symbols show the results from the fits shown in
    Figure~\ref{fig:000-400-fit}, while small symbols show results for
    fits using \(\Delta\theta = \ang{60}\) instead of \ang{75}. In panel
    \textit{a}, two slightly different \(\Pi'\) values are shown for the
    outer bow, since the fit does not converge to a single value (see
    Fig.~\ref{fig:000-400-fit}\textit{c} and
    App.~\ref{app:rcurv-empirical}).}
  \label{fig:000-400-planitude-alatude}
\end{figure*}

\begin{figure*}
  \begin{tabular}{p{0.47\linewidth} p{0.47\linewidth}}
    (a) & (b) \\
    \includegraphics[width=\linewidth]{figs/000-400-Pi-vs-Dtheta}
    & \includegraphics[width=\linewidth]{figs/069-601-Pi-vs-Dtheta}
  \end{tabular}
  \caption[]{Variation of fitted planitude, \(\Pi'\), as a function of
    the parameter \(\Delta\theta\), which controls how close a point must be to
    the axis in order to be included in the circle
    fit. (\textit{a})~M42 000-400, (\textit{b}) M42 069-601.  For the
    three traced bows (inner, outer, and ridge) of each object, the
    symbol key lists the mean and standard deviation of \(\Pi'\),
    calculated over the ``stable'' range \(\Delta\theta = \ang{60}\) to
    \ang{80}, which is indicated by light gray shading.}
  \label{fig:000-400-Delta-theta}
\end{figure*}

@@

As an example of measuring the projected planitude and alatude of real
bow shocks, we present an analysis of M42~000-400 and M42~069-601,
which are two H\(\alpha\) emission arcs \citep{Bally:2000a,
  Gutierrez-Soto:2015a} associated with proplyds\footnote{%
  The coordinate designation (see \citealp{ODell:1994a} for an
  explanation of the nomenclature) of 000-400 is very imprecise in
  right ascension, but we use it here for consistency with previous
  papers. The associated proplyd is listed with the more correct
  designation 4596-400 in catalogs such as \citet{Ricci:2008a}.  } %
in the west of the Orion Nebula (M42).  An image of one of these arcs
(M42~069-601) was used in the illustration of bow shock terminology in
Figure~\ref{fig:bow-terminology}.

We consider three different tracings of the bow shape (see
Fig.~\ref{fig:000-400-fit}): the peak of the emission arc (``ridge''),
and its inner and outer edges.  In all three cases, we placed by eye
the points that define the bow, using SAOImage~DS9 in a similar
fashion to in Figure~\ref{fig:meyer-trace}, and guided by the image
contours.\footnote{%
  In the case of the ``ridge'' method at least, it is possible to
  automate this step, which we will discuss in detail in a following
  paper. } %

We determine the planitude and alatude by fitting a circle to the
traced points within \(\pm \Delta\theta = \ang{75}\) bow axis, using the
iterative algorithm described in Appendix~\ref{app:rcurv-empirical}.
The fitted circle, when combined with the position of the central
source, yields the orientation of the bow axis, together with the apex
distance, \(R_0\), radius of curvature, \(R_\C\), and two
perpendicular radii (one for each wing), \(R_{90+}\) and \(R_{90-}\).
These are all indicated on the panels of Figure~\ref{fig:000-400-fit}
by light-colored lines.\footnote{%
  For conciseness, we drop the prime symbol from the radii, both in
  this section and in Appendix~\ref{app:rcurv-empirical}, even though
  they are all projected quantities.} %
The projected planitude and alatude then follow as \(\Pi' = R_\C/R_0\),
\(\Lambda_+ = R_{90+}/R_0\), \(\Lambda_- = R_{90-}/R_0\), which are shown in
Figure~\ref{fig:000-400-planitude-alatude}.

The planitude is found to have a moderate dependence on the choice of
\(\Delta\theta\), as shown in Figure~\ref{fig:000-400-Delta-theta}, where it can
be seen that, although the values of \(\Pi'\) are relatively stable for
\(\Delta\theta \ge \ang{60}\) (light gray shaded region), they can show much
larger variations for \(\Delta\theta < \ang{60}\).  The fact that the radius of
curvature is defined at a point (the projected apex) might seem to
argue for making \(\Delta\theta\) as small as possible, but that would lead to
circle fits that were extremely sensitive to the exact positions of
the few points included in the fit.  A reliable fit requires 4 or more
points, ideally spanning a total separation that is a substantial
fraction of \(R_\C\), which would argue for \(\Delta\theta\) larger than about
\(\Pi'/2\) radians, or \ang{60}--\ang{90}.  On the other hand, if
\(\Delta\theta \ge \ang{90}\) then the planitude and alatude would no longer be
independent since the bow would be forced to lie on the ``sphere''
line, \(\Lambda' = (2 \Pi' - 1)^{1/2}\) (see \S~\ref{sec:conic}).  Balancing
these two concerns suggests an optimal \(\Delta\theta = \ang{75}\), which is
shown in Figure~\ref{fig:000-400-fit}, whereas in
Figure~\ref{fig:000-400-planitude-alatude} we show results for both
\ang{75} (thick lines) and \ang{60} (thin lines).

\begin{table}
  \caption[x]{Fitted planitude and alatude for M42 bows}
  \label{tab:m42-fit}
  \centering
  \begin{tabular}{ll lll}
    \toprule
    Source & Bow & \(\Pi'\) & \(\Lambda'\)& \(\Delta\Lambda'\) \\
    \midrule
    000-400 & Inner & \(2.87 \pm 0.19\) & \(2.40 \pm 0.04\) & \(0.35 \pm 0.08\) \\
           &  Ridge & \(3.62 \pm 0.38\) & \(2.73 \pm 0.02\) & \(0.51 \pm 0.10\) \\
           & Outer & \(2.51 \pm 0.24\) & \(2.46 \pm 0.02\) & \(-0.06 \pm 0.02\) \\[\smallskipamount]
           & All & \(3.00 \pm 0.54\) & \(2.53 \pm 0.15\) & \(0.16 \pm 0.16\) \\[\bigskipamount]
    069-601 & Inner & \(3.81 \pm 0.23\) & \(2.90 \pm 0.01\) & \(0.44 \pm 0.05\) \\
           & Ridge & \(2.43 \pm 0.13\) & \(2.17 \pm 0.01\) & \(0.07 \pm 0.02\) \\
           & Outer & \(2.22 \pm 0.15\) & \(2.11 \pm 0.01\) & \(0.01 \pm 0.01\) \\[\smallskipamount]
           & All & \(2.82 \pm 0.75\) & \(2.39 \pm 0.37\) & \(0.17 \pm 0.20\)\\
    \bottomrule
  \end{tabular}
\end{table}

\(\Pi' = 2.9 \pm 0.2 \pm 0.3\)

The principal hindrance to an accurate tracing of the bow shape in
this object is the strong and variable nebular background emission,
which particularly affects the far wing on the right hand side of the
image, where a bright background filament is superimposed on the bow.

These give an idea of the systematic uncertainties and sensitivity to
subjective choices,
